---
title: "mall — live examples"
subtitle: "Automatic NLP with LLMs Using {mall}"
format: html
toc: true
toc-depth: 2
engine: knitr
execute:
  eval: true
  echo: true
---

## Setup

```{r}
#| label: setup
library(mall)
library(dplyr)
```

### Local model via Ollama

```{r}
#| label: llm-use-local
llm_use("ollama", "llama3.2", seed = 100, .cache = "_mall_cache")
```

### Remote model via ellmer (alternative)

To use a remote model instead, swap the `llm_use()` call above for one of these:

```{r}
#| eval: false
library(ellmer)

# OpenAI
llm_use(chat_openai(model = "gpt-4o"))

# Anthropic
llm_use(chat_anthropic(model = "claude-haiku-4-5-20251001"))

# Google Gemini
llm_use(chat_google_gemini())
```

---

## The `reviews` dataset

`mall` ships with a small built-in dataset — 9 product reviews.

```{r}
#| label: reviews-data
data(reviews)
reviews
```

---

## Sentiment analysis

`llm_sentiment()` classifies each row as **positive**, **negative**, or **neutral**.

```{r}
#| label: sentiment-default
reviews |>
  llm_sentiment(review)
```

You can specify your own labels:

```{r}
#| label: sentiment-custom-labels
reviews |>
  llm_sentiment(review, options = c("happy", "sad", "meh"))
```

---

## Summarize

`llm_summarize()` condenses each review to a target number of words.

```{r}
#| label: summarize
reviews |>
  llm_summarize(review, max_words = 5)
```

---

## Classify

`llm_classify()` assigns each row to one of your predefined categories.

```{r}
#| label: classify
reviews |>
  llm_classify(review, labels = c("appliance", "computer", "tv"))
```

---

## Extract entities

`llm_extract()` pulls a specific piece of information from each row.

```{r}
#| label: extract-product
reviews |>
  llm_extract(review, "product name")
```

Extract multiple things by calling `llm_extract()` more than once:

```{r}
#| label: extract-multiple
reviews |>
  llm_extract(review, "product name") |>
  llm_extract(review, "main complaint or compliment")
```

---

## Verify

`llm_verify()` answers a yes/no question about each row — returns **1** (yes) or **0** (no).

```{r}
#| label: verify
reviews |>
  llm_verify(review, "The customer would recommend this product")
```

Combine with `filter()` to find rows that match a condition:

```{r}
#| label: verify-filter
reviews |>
  llm_verify(review, "The customer would recommend this product") |>
  filter(.verify == 1)
```

---

## Translate

`llm_translate()` translates each row into any target language — no language-specific
model required.

```{r}
#| label: translate-to-spanish
reviews |>
  llm_translate(review, "spanish")
```

---

## Multilingual workflow

A realistic end-to-end example: detect the language, keep non-English rows,
translate them to English, then run sentiment analysis.

For this demo we'll build a small mixed-language dataset:

```{r}
#| label: multilingual-data
multilingual <- tibble(
  review = c(
    "This product is amazing, I love it!",
    "Este producto es terrible, muy decepcionante.",
    "Je suis très satisfait de cet achat, excellent rapport qualité-prix.",
    "Das Produkt ist okay, aber nichts Besonderes.",
    "Worst purchase I have ever made, do not buy this."
  )
)

multilingual
```

Step 1 — detect language:

```{r}
#| label: multilingual-detect
multilingual |>
  llm_extract(review, "language (e.g. english, spanish, french, german)")
```

Step 2 — full pipeline (detect → filter non-English → translate → sentiment):

```{r}
#| label: multilingual-pipeline
multilingual |>
  llm_extract(review, "language") |>
  filter(.extract != "english") |>
  llm_translate(review, "english") |>
  llm_sentiment(.translation)
```

---

## Custom prompts

`llm_custom()` lets you write any prompt. Use it for tasks that don't fit the
built-in functions.

```{r}
#| label: custom-feature-mention
reviews |>
  llm_custom(
    review,
    "Reply with only 'yes' or 'no': Does this review mention a specific product feature?"
  )
```

```{r}
#| label: custom-score
reviews |>
  llm_custom(
    review,
    "Rate the urgency of this customer feedback from 1 (low) to 5 (high). Reply with the number only."
  )
```

---

## Vector functions

Every `llm_*()` function has an `llm_vec_*()` counterpart for working on a single string.

```{r}
#| label: vec-sentiment
llm_vec_sentiment("I absolutely love this product!")
```

```{r}
#| label: vec-translate
llm_vec_translate("Buenos días a todas", "english")
```

```{r}
#| label: vec-extract
llm_vec_extract("Please call me back at 555-1234 before noon.", "phone number")
```

```{r}
#| label: vec-summarize
llm_vec_summarize(
  "The battery life on this phone is outstanding. I charged it on Monday and it is now Thursday and I still have 40% remaining. Highly recommend.",
  max_words = 8
)
```

---

## Chaining everything together

A single pipeline combining multiple `mall` functions:

```{r}
#| label: full-pipeline
reviews |>
  llm_classify(review, labels = c("appliance", "computer", "tv")) |>
  llm_sentiment(review) |>
  llm_summarize(review, max_words = 6) |>
  select(review, .classify, .sentiment, .summary)
```

---

## Caching

Results are cached by default in `_mall_cache/`. Re-running the same text
with the same model returns the cached result instantly — no API call needed.

```{r}
#| eval: false
# Use a custom cache folder
llm_use(.cache = "_my_cache")

# Disable caching entirely
llm_use(.cache = "")

# Re-enable with default folder
llm_use(.cache = "_mall_cache")
```

---

## Session info

```{r}
#| label: session-info
sessionInfo()
```
