---
title: "Automatic NLP with LLMs Using `mall`"
format: html
toc: true
toc-depth: 2
engine: knitr
execute:
  eval: true
  echo: true
---

## Setup

```{r}
library(mall)
library(dplyr)
```

### Local model via Ollama

```{r}
llm_use("ollama", "llama3.2")
```

```{r}
llm_use("ollama", "llama3.2", .cache = "_mall_cache")
```

```{r}
llm_use("ollama", "llama3.2", .cache = "_mall_cache", seed = 100)
```


## The `reviews` dataset

`mall` ships with a small built-in dataset with 9 product reviews.

```{r}
data(reviews)
reviews
```


## Sentiment analysis

`llm_sentiment()` classifies each row as **positive**, **negative**, or **neutral**.

```{r}
reviews |>
  llm_sentiment(review)
```

You can specify your own labels:

```{r}
reviews |>
  llm_sentiment(review, options = c("happy", "sad", "meh"))
```


## Summarize

`llm_summarize()` condenses each review to a target number of words.

```{r}
reviews |>
  llm_summarize(review, max_words = 5)
```



## Classify

`llm_classify()` assigns each row to one of your predefined categories.

```{r}
reviews |>
  llm_classify(review, labels = c("appliance", "computer", "tv"))
```



## Extract entities

`llm_extract()` pulls a specific piece of information from each row.

```{r}
reviews |>
  llm_extract(review, "product name")
```

Extract multiple things by calling `llm_extract()` more than once:

```{r}
reviews |>
  llm_extract(review, "product name", pred_name = "product_name") |>
  llm_extract(review, "main complaint or compliment", pred_name = "reason")
```



## Verify

`llm_verify()` answers a yes/no question about each row returns **1** (yes) or **0** (no).

```{r}
reviews |>
  llm_verify(review, "The customer would recommend this product")
```

Combine with `filter()` to find rows that match a condition:

```{r}
reviews |>
  llm_verify(review, "The customer would recommend this product") |>
  filter(.verify == 1)
```



## Translate

`llm_translate()` translates each row into any target language that the LLM 
can understand

```{r}
reviews |>
  llm_translate(review, "spanish")
```

```{r}
reviews |>
  llm_translate(review, "french")
```

## Multilingual workflow

A realistic end-to-end example: detect the language, keep non-English rows,
translate them to English, then run sentiment analysis.

For this demo we'll build a small mixed-language dataset:

```{r}
multilingual <- tibble(
  review = c(
    "This product is amazing, I love it!",
    "Este producto es terrible, muy decepcionante.",
    "Je suis très satisfait de cet achat, excellent rapport qualité-prix.",
    "Das Produkt ist okay, aber nichts Besonderes.",
    "Worst purchase I have ever made, do not buy this."
  )
)

multilingual
```

Step 1detect language:

```{r}
multilingual |>
  llm_extract(review, "language (e.g. english, spanish, french, german)")
```

Step 2, full pipeline (detect → filter non-English → translate → sentiment):

```{r}
multilingual |>
  llm_extract(review, "language") |>
  filter(.extract != "english") |>
  llm_translate(review, "english") |>
  llm_sentiment(.translation)
```



## Custom prompts

`llm_custom()` lets you write any prompt. Use it for tasks that don't fit the 
built-in functions.

Not documented, use `{x}` to place the current row's data.

```{r}
my_prompt <- paste(
  "Answer a question.",
  "Return only the answer, no explanation.",
  "Choose the best answer between: yes or no",
  "Answer this about the following text, is this a happy customer?: {x}"
)

reviews |>
  llm_custom(review, my_prompt)

```


If you need to ensure the reponses are within a specified set

```{r}
reviews |>
  llm_custom(review, my_prompt, valid_resps = c("Yes", "Maybe"))
```

## Vector functions

Every `llm_*()` function has an `llm_vec_*()` counterpart for working on a single string.

```{r}
llm_vec_sentiment("I absolutely love this product!")
```

```{r}
llm_vec_translate("Buenos días a todas", "english")
```

```{r}
llm_vec_extract("Please call me back at 555-1234 before noon.", "phone number")
```

```{r}
llm_vec_summarize(
  "The battery life on this phone is outstanding. I charged it on Monday and it is now Thursday and I still have 40% remaining. Highly recommend.",
  max_words = 8
)
```


